Data exploration, quality control and statistical analysis of ChIP-exo experiments

### 2016-05-09 rmarkdown/paper.Rnw
fixed version of paper

Using the regions and summary statistics we obtained a set of
candidate sites for each replicate, the total number of candidate
sites is shown in Figure \ref{fig:enrich}B. For this sites, we
extracted the sequence and searched for the FoxA1 motif using
\textbf{FIMO} \cite{fimo}. Figure \ref{fig:enrich}C shows the fraction
of candidate site where a motif was found. The first replicate
outperforms the rest in both number of candidate sites and proportion
of sites with motif; on the other hand we can see that the second
replicate candidate sites are more likely to contain the motif
sequence, but there more candidate sites in the third replicate than
in the second. Finally, Figure \ref{fig:enrich}D shows the average
profiles around the coordinate where the motif was found in the
candidate sites: We can observe that the signal is more delimited for
the first and third replicates than the second one, but overall the
signal pattern is reproducible in the three ChIP-exo
experiments. Figures \ref{fig:enrich}E, \ref{fig:enrich}F and
\ref{fig:enrich}G show the purity of the motif found by visualizing
the sequence to which the motif was matched. All these observations
together shows the following findings: First, observing a defined
decreasing trend in $\mbox{URCR}$ as $\mbox{ARC}$ increases implies
than the sample is more enriched, therefore a higher amount of
candidate sites. Second, for experiments that are not deeply
sequenced, low $\mbox{URCR}$ values represent low complexity which
decreases the amount of sites of motif found but it doesn't modify the
read distribution around the motif. Finally, libraries with higher
complexity returns sequences with a better motif match.

% To verify this statement, we
% extracted the sequences around high confidence binding events and look
% for the FoxA1 motif using FIMO \cite{fimo}. Figure \ref{fig:enrich}B
% shows the number of candidate regions, which shows that the first
% replicate is being allocated into more enriched regions than the other
% ones. Figure \ref{fig:enrich}C shows that for the first and third
% replicates, the FoxA1 motif is being detected in roughly the same
% proportion of sites, and finally in \ref{fig:enrich}D we observe that
% the first and third replicate can detect the FoxA1 motif with the same
% significance, while the second replicate does not.
\color{blue}

outline

\begin{enumerate}
\item \st{Available data, the motivation of this part is two fold:}
 \begin{itemize}
 \item \st{we want to list all data sets we obtained}
 \item \st{we want to emphasize that for the sig70 datasets we have a
   unique design that allows us to compare chip exo with SE and PE
   chipseq , the later one for the first time}
 \end{itemize}
\item \st{comparison with chip seq} %? with and without dpeak ?}
\item \st{only in methods. list the align algorithm used}
\item \st{chip seq qc metrics}
\item \st{chip exo qc pipeline, description and justification of the steps}
\begin{itemize}
\item need to justify the sampling step. for this we may use the depth
  table and add the factor to normalize one depth to another. explain
  the confounding of pcr amplification and the reduction of the number
  of possible mappable positions.
\item \st{case study with FoxA1, analysis of enrichment vs library
  complexity and analysis and the strand imbalance tools. in methods
  we need to add which tests did we perform to show that the pipeline
  works. how do we expect from the diagnostics and how to interpret
  the plots.}
\item analysis of why the pipeline works. in methods need to specify
  how and in results need to specify why it works, and what are the
  conclusions from it
\end{itemize}
\item high resolution analysis with chip exo and comparison with PE
  chip seq
  \begin{itemize}
  \item justify what do we used dpeak (in methods need to list the
    other methods used). perhaps we can add gof plots in
    supplement. we also want to compare with the same method,
    therefore we can use the fact that dpeak have a unified model for
    PE and SE dpeak
  \item after comparing with other methods we can compare with PE and
    SE ChIP-seq, this comparison gives also resolution and sensitivity
  \item compare samples under same budget of reads
  \end{itemize} 
\end{enumerate}

\color{black}

@@

The strand imbalance assessment is based in the observation that the
enriched regions usually are composed of a higher quantity of reads,
therefore we examined the FSR (defined as the ratio of number of
forward stranded reads divided by the total number of reads in a given
region) as the regions with lower depth are being filtered out. This
indicator is of particular importance, as it evaluates the ``peak
pair'' assumption that the original ChIP-exo paper suggested and
multiple ChIP-exo data analysis methods rely on. For every ChIP-exo
experiment, we calculated the global FSR and noticed that for all
experiments is roughly $0.5$, which means there are roughly the same
amount of reads in both strands.

In order to asses the strand imbalance we created the visualization
shown in Figure \ref{fig:strand}: Figure \ref{fig:strand}A presents
the FSR's behavior as the lower depth regions are being filtered out,
while Figure \ref{fig:strand}B) shows which percentage of the regions
are composed by reads in both strand or only one (forward or
backward). In a good data set, it would be expected that all quantiles
shown to be quickly converging towards the median (in panel A) or the
regions composed of reads in one strand being made of few fragments
(in panel B). For each replicate, we divided the partitioned regions
by asking whether they overlap a set of high quality ChIP-exo peaks,
and then we tested (using the Wilcoxon rank sum test over the
imbalance index defined in the methods section) if the strand
imbalance's distribution is the same for both classes. For regions
composed by a higher amount of reads, it is harder to distinguish
their peaks by considering only the strand imbalance, hence in a
better quality ChIP-exo experiment it is easier to distinguish
enriched regions by the amount of reads in both strands. Similarly, we
may consider that the strands of reads for background of a ChIP-exo
experiment is more unbalanced than those for the enriched regions. In
conclusion, Figure \ref{fig:strand} shows that the global FSR does not
represent the experiment's local strand imbalance, hence the ``peak
pair'' assumption does not hold locally in ChIP-exo data.

\textbf{Analysis of ChIP-nexus data with the QC pipeline}

We applied the QC pipeline to the ChIP-nexus experiments and found
that the overall pattern seen in the visualizations is similar to a
high quality ChIP-exo experiment. Figure \ref{fig:nexus}A illustrates
a typical pattern of the ARC vs. URCR plot with ChIP-nexus data: Both
arms seems to be more narrow and therefore the separation between them
seems to be more noticeable, this seems reasonable since libraries
generated by ChIP-nexus usually yields higher complexity than
libraries generated by ChIP-exo. Additionally, while in ChIP-exo
experiments is common to observe regions formed by undigested
fragments in only one strand, we can observe that in ChIP-nexus
experiment the occurrence of this type of regions is reduced. Figure
\ref{fig:nexus}B shows that in a ChIP-nexus experiment the FSR
distribution approximates the median more quickly than in a ChIP-exo
experiment and Figure \ref{fig:nexus}C we observe that most of these
single-stranded regions are formed by few fragments, which means that
good ChIP-nexus experiment are less likely to show very imbalanced
regions.

\subsubsection{Evaluating ChIP-exo and ChIP-nexus data}

<<all_param,include = FALSE,echo = FALSE,eval = TRUE>>=

  Ntimes <- prettyNum(1e4,big.mark = ",")
  Msamp <- prettyNum(1e3,big.mark = ",")
  thresh1 <- 10

We applied the QC pipeline to all ChIP-exo and ChIP-nexus datasets
listed in Tables \ref{tab:qc} and \ref{tab:nexus}, that way we
partitioned the reads of each experiment into the undigested regions
by keeping the regions formed by at least one fragment. Figure
\ref{fig:enrich_all}A shows compares the slope of the following model:

\begin{align}
  \mbox{depth} = \beta \times \text{(\# of unique positions)} +\epsilon  \nonumber
\end{align}

when sampling \Sexpr{Msamp} regions and repeating this process
\Sexpr{Ntimes} times for every ChIP-exo and ChIP-nexus experiment. The
slope can be interpreted as the change in depth as number of unique
position varies. This parameter measures the library complexity on a
ChIP-exo experiment, since when it is high it means that several
fragments are being aligned to the same position in the genome.

Figures \ref{fig:enrich_all}B and \ref{fig:enrich_all}C shows the
parameters of the following model:

\begin{align}
  \mbox{URCR} = \frac{\kappa}{\mbox{ARC}} + \eta + \epsilon \nonumber
\end{align}

considering only the regions composed by 10 or more aligned reads,
that way the model reflects the enriched arm of the ARC vs. URCR
visualization. The intercept $\eta$ can be interpreted as the
theoretical $\mbox{URCR}$ for relatively high depth regions and the
slope $\kappa$ can be interpreted as the decay in $\mbox{URCR}$ as the
$\mbox{ARC}$ increases. That way experiments with low complexity are
showing low values for both parameters. We considered a linear model
between the estimated parameters for each experiment and it's
respective depth for which we were able to asses that there is not
significant relationship between the estimated parameters and the
interaction between depth and organism for each experiment. Hence,
these parameters may be used as indicators of the overall quality of
ChIP-exo experiment, but as with the SCC it is recommendable to also
visualize the SCC curve.

\subsection{High Resolution Binding Site Identification with dPeak and
  ChIP-exo}

\subsubsection{Comparison with ChIP-Seq data using dPeak}

@@

Figure \ref{fig:reso_all} shows comparisons among ChIP-exo, PE
ChIP-Seq and SE ChIP-Seq. We considered the RegulonDB data as ground
truth, since those are the most recent annotation on \emph{Escherichia
  Coli}. A RegulonDB annotation (Salgado et al, 2012 \cite{regulondb})
was considered to be identified if the distance from the closest dPeak
binding site estimate was less that or equal to $\Sexpr{id_dist}$
bp. That way, the sensitivity is defined as the proportion of
RegulonDB annotations identified in a peak and the resolution is
defined as the minimum distance between a RegulonDB annotation and the
closest dPeak binding site estimate. Figure \ref{fig:reso_all}A shows
that the sensitivity increases as the mean distance between binding
events increases. When the average distance is greater than 200 bp,
dPeak identifies more than the 75\% of the binding events in each
peak, this is intuitive as the mean ChIP-Seq's fragment length is
shorter than 200 bp, hence the read does not contribute to more than
one binding event \cite{dpeak}. When the binding events in a peak are
closer to each other, both ChIP-exo and PE ChIP-Seq are comparable, as
the distance increases ChIP-exo identifies a higher proportion of the
RegulonDB annotations; additionally SE ChIP-Seq is significantly less
sensitive than both ChIP-exo and PE ChIP-Seq. Figure
\ref{fig:reso_all}B shows that ChIP-exo and PE ChIP-Seq are comparable
in resolution, while both protocols significantly outperform SE
ChIP-Seq.

\subsubsection{Systematic comparison of ChIP-Seq vs ChIP-exo under
  varying sequencing depth}

Previously, ChIP-exo and SE ChIP-Seq have been compared only at a
fixed depth level in the literature, while they did not include PE
ChIP-Seq as well either. In order to address this limitation in
previous studies, we sampled a fixed amount of reads for each of the
ChIP-exo, PE ChIP-Seq and SE ChIP-Seq datasets of the $\sig$ samples
($N$ reads for both ChIP-exo experiment and $N/2$ or $N$ pairs for PE
ChIP-Seq to assume more realistic situation of a fixed cost). For each
sampled dataset we applied our lower-to-higher resolution pipeline by
calling peaks with MOSAiCS \cite{mosaics} and then deconvolving the
binding events by using dPeak \cite{dpeak}. For the ChIP-exo datasets
we called peaks by using the GC-content and mappability models with
MOSAiCS, since it's background is usually composed of scattered reads
across the genome; and for the ChIP-Seq datasets we used their
respective Input samples. Additionally, it is worth noting that for PE
ChIP-Seq we sampled both ends of the fragment, hence for each
sequencing depth we are sampling the half amount of pairs for PE
ChIP-Seq than for ChIP-exo or SE ChIP-Seq.

Figures \ref{fig:design} shows the behavior of each data type in
$\sig$ experiment under aerobic condition when comparable number of
reads is used for all of ChIP-exo, SE ChIP-Seq and PE ChIP-Seq. In
Figure \ref{fig:design}A we show the number of candidate regions
defined as the number of regions where a binding event was identified
in a collection of high quality peaks; in Figure \ref{fig:design}B we
depict the number of binding events; in Figure \ref{fig:design}C we
show the number of identified targets, where a RegulonDB was
considered identified if a binding event was identified in a 15 bp
vicinity of it; and finally Figure \ref{fig:design}D show the
resolution defined as the distance from a RegulonDB annotation to the
closest dPeak prediction. It is remarkable that even when the number
of candidate peaks or the number of predicted events is lower for
ChIP-exo, it outperforms both PE and SE ChIP-Seq in number of
identified targets and resolution.

Figure \ref{fig:saturation_rif} shows an analogous analysis but using
the $\sig$ replicates with and without rif treatment. The left, middle
and right columns show the fixed depth against the number of predicted
events, identified targets and resolution being compared at a fixed
depth level. The behavior of this quantities seems to be opposite to
the one as in Figure \ref{fig:design}, hence we used the ChIP-exo QC
pipeline in the fixed depth ChIP-exo experiments. Figure
\ref{fig:exoQC_sat_aero} shows scatter plots of ARC vs URCR for
several fixed sample sizes, as the fixed depth increases the two arm
pattern becomes more distinctive while for lower depth, it seems that
the majority of the sampled reads were aligned to enriched regions. On
the other hand in Figures \ref{fig:exoQC_sat1} to
\ref{fig:exoQC_sat4}, we used the ChIP-exo QC pipeline on the samples
that are outperformed by PE and SE ChIP-Seq. For a fixed low depth, we
can see that the majority of the reads are being aligned to
non-enriched regions since the vertical arm seems to be stronger for
all 2 conditions and replicates; for higher depth we can see low URCR
values being predominant, which indicates that the majority of the
regions being formed by few positions with a higher read
concentration. In low complexity regions, the reads are being aligned
to fewer positions but there is no control over the amount of reads
mapped. Hence, those regions are more likely to being strand-imbalance
which in turn may bias the binding site estimate and therefore
decrease the number of identified targets or increase an experiments
resolution.

% In supplement figures \ref{fig:exoQC_sat1} to \ref{fig:exoQC_sat4}, we
% repeat the same analysis for both replicates under the two rif
% conditions.

\subsubsection{dPeak outperforms competing methods in discovering closely
  spaced binding events from ChIP-exo and ChIP-Seq data}

Figure \ref{fig:methods_comp} compares the resolution defined as the
minimum distance between a RegulonDB annotation and a binding site
predicted by either Peakzilla \cite{peakzilla}, MACE \cite{mace}, GEM
\cite{gem} or dPeak \cite{dpeak}. In a good dataset such as both of
the ChIP-exo experiments under aerobic (panel 931 and 933) conditions,
all the methods are comparable in resolution, and dPeak slightly
outperforms the rest. On the other hand, for experiments dPeak
identified binding sites with higher resolution than Mace and
Peakzilla, but with low resolution than Gem. This may be due that the
fact that Gem uses sequence information in addition to the aligned
$5\prime$ end counts that dPeak uses. 
 \footnote{For here we may
  probably use only 933 as part of the main article and keep the rest
  for the supplement}

@@

  fdr <- 0.01
  topM <- 300

@@

@@

\begin{figure}[h!]
  \centering  
  \includegraphics[width = .9\textwidth,page = 1]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \newline
  \includegraphics[width = .9\textwidth,page = 2]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \includegraphics[width = .4\textwidth]{../figs/for_paper/Carroll_FSR_depth_VS_pvalWilcoxon.pdf}
  \caption{Strand imbalance QC plots for the same data as in Figure
    \ref{fig:enrich}A. A) FSR distribution quantiles as the lower
    depth regions are being filtered out, all quantiles approach to
    the median as the lower bound increases. B) Stacked histogram with
    the proportion of regions that are formed by two strands or only
    one, in a good sample the single-stranded regions are going to be
    filtered out quickly as in the middle row. C)
    $-\log_{10}(\text{p.value})$ of testing if the imbalance
    distributions differs when ChIP-exo regions overlap their peaks.}
  \label{fig:strand}
\end{figure}

>>>>

Considering the regions and summary statistics we obtained a set of
candidate sites for each replicate, the total number of candidate
sites is shown in Figure \ref{fig:enrich}B. For this sites, we
extracted the sequence and searched for the FoxA1 motif using
\textbf{FIMO} \cite{fimo}. Figure \ref{fig:enrich}C shows the fraction
of candidate site where a motif was found. The first replicate
outperforms the rest in both number of candidate sites and proportion
of sites with motif; on the other hand we can see that the second
replicate candidate sites are more likely to contain the motif
sequence, but there more candidate sites in the third replicate than
in the second. Finally, Figure \ref{fig:enrich}D shows the average
profiles around the coordinate where the motif was found in the
candidate sites: We can observe that the signal is more delimited for
the first and third replicates than the second one, but overall the
signal pattern is reproducible in the three ChIP-exo
experiments. Figures \ref{fig:enrich}E, \ref{fig:enrich}F and
\ref{fig:enrich}G show the purity of the motif found by visualizing
the sequence to which the motif was matched. All these observations
together shows the following findings: First, observing a defined
decreasing trend in $\mbox{URCR}$ as $\mbox{ARC}$ increases implies
than the sample is more enriched, therefore a higher amount of
candidate sites. Second, for experiments that are not deeply
sequenced, low $\mbox{URCR}$ values represent low complexity which
decreases the amount of sites of motif found but it doesn't modify the
read distribution around the motif. Finally, libraries with higher
complexity returns sequences with a better motif match.

% To verify this statement, we
% extracted the sequences around high confidence binding events and look
% for the FoxA1 motif using FIMO \cite{fimo}. Figure \ref{fig:enrich}B
% shows the number of candidate regions, which shows that the first
% replicate is being allocated into more enriched regions than the other
% ones. Figure \ref{fig:enrich}C shows that for the first and third
% replicates, the FoxA1 motif is being detected in roughly the same
% proportion of sites, and finally in \ref{fig:enrich}D we observe that
% the first and third replicate can detect the FoxA1 motif with the same
% significance, while the second replicate does not.

@@

% Using the Fox A1 in mouse liver cell lines generated by Serandour et
% al., 2013\nocite{exoillumina}

The strand imbalance assessment is based in the observation that the
enriched regions usually are composed of a higher quantity of reads,
therefore we examined the FSR as the regions with lower depth are
being filtered out. This indicator is of particular importance, as it
evaluates the ``peak pair'' assumption that the original ChIP-exo
paper suggested and multiple ChIP-exo data analysis methods rely
on. For every ChIP-exo experiment, we calculated the global FSR and
noticed that for all experiments is roughly $0.5$, which means there
are approximately the same amount of reads in both strands.

In order to study the strand imbalance of the FoxA1 factor ChIP-exo
experiments in mouse liver generated by Serandour et al.,
2013\nocite{exoillumina} we considered the visualizations shown in
Figure \ref{fig:strand}: Figure \ref{fig:strand}A presents the FSR's
behavior as the lower depth regions are being filtered out, while
Figure \ref{fig:strand}B) shows the proportion of regions that are
composed by reads from both strands against the regions formed by
reads in exclusively one strand. In a high quality dataset, it is
expected for all quantiles to quickly converging towards the median
(in panel A) or the regions formed by reads in one strand (either
forward or backward) to be formed by few fragments (in panel
B).

Additionally, it is worth mentioning that the background in a ChIP-exo
experiment is expected to be formed by the undigested ChIP fragments
that are not bound to a protein, hence it is for the reads in both
strand to be imbalanced in a background region. Figure
\ref{fig:strand}C compares the strand imbalance when ChIP-exo regions
overlap with high quality peaks. It is noticeable that for regions
composed by a large amount of reads, it is harder to distinguish their
peaks by considering only the strand imbalance, hence in a high
quality ChIP-exo experiment the background is expected to show a
higher imbalance that the enriched regions. In conclusion, Figure
\ref{fig:strand} shows that the global FSR does not accurately
represent a ChIP-exo experiment's strand imbalance locally, hence the
``peak pair'' assumption does not completely hold in every ChIP-exo
enriched region.

% \textbf{Analysis of ChIP-nexus data with the QC pipeline}

% We applied the QC pipeline to the ChIP-nexus experiments and found
% that the overall pattern seen in the visualizations is similar to a
% high quality ChIP-exo experiment. Figure \ref{fig:nexus}A illustrates
% a typical pattern of the ARC vs. URCR plot with ChIP-nexus data: Both
% arms seems to be more narrow and therefore the separation between them
% seems to be more noticeable, this seems reasonable since libraries
% generated by ChIP-nexus usually yields higher complexity than
% libraries generated by ChIP-exo. Additionally, while in ChIP-exo
% experiments is common to observe regions formed by undigested
% fragments in only one strand, we can observe that in ChIP-nexus
% experiment the occurrence of this type of regions is reduced. Figure
% \ref{fig:nexus}B shows that in a ChIP-nexus experiment the FSR
% distribution approximates the median more quickly than in a ChIP-exo
% experiment and Figure \ref{fig:nexus}C we observe that most of these
% single-stranded regions are formed by few fragments, which means that
% good ChIP-nexus experiment are less likely to show very imbalanced
% regions.

\subsubsection*{Evaluation of ChIP-exo and ChIP-nexus data with QC
  pipeline}

We evaluated every experiment listed in Tables \ref{tab:qc_sig} and
\ref{tab:qc} with the QC pipeline. For each experiment, we explored
the relationship between depth, width and the number of unique
positions to summarize the shape obtained in the $\mbox{ARC}$
vs. $\mbox{URCR}$ visualization. Figure \ref{fig:eval}A shows the
overall change in depth as the number of unique positions varies for
the gathered experiment in eukaryotic genomes. This parameter can be
interpreted as the inverse of $\gamma$ in equation (\ref{mod}), which
can is also the experiment's large depth $\mbox{URCR}$. In a high
quality ChIP-exo experiment, it is expected in the $\mbox{ARC}$
vs. $\mbox{URCR}$ plot to observe two arms, one with low $\mbox{ARC}$
and varying $\mbox{URCR}$ and another that with a decreasing
$\mbox{URCR}$ as $\mbox{ARC}$ increases that stabilizes in
$\gamma$. When the ChIP-exo experiment is not deeply sequenced, high
values in Figure \ref{fig:eval} indicate that the library complexity
from the experiment is low. On the other hand, low values corresponds
to higher quality ChIP-exo experiment.

Figure \ref{fig:eval}B shows the average read coverage bias when the
experiment's sequencing depth is large/ Under perfect digestion
setting, most of the reads aligned to binding regions are going to be
accumulated around a binding event, which suggest to be unlikely to
observe as the sequencing depth increases to observe a reads being
aligned to another position that was not previously covered. Low
quality ChIP-exo experiment exhibit more scattered reads in both
strands across the genome; therefore it is more likely to observe
consider reads that align to position in the genome that were not
previously covered. We can compare the second replicate from the human
ChIP-nexus experiment against both second and third replicates from
the TBP factor experiment in K562 cell lines. The sequencing depth of
this three experiment is comparable but the large depth $\mbox{ARC}$'s
bias is considerably higher for both ChIP-exo experiments.

The interaction between these two parameters represents the quality of
a ChIP-exo and ChIP-nexus experiment. When either the adjusted
$\mbox{ARC}$ or the $\mbox{ARC}$'s bias is large, then it is suggested
to sub-sample reads from the experiment and then evaluate the
sub-sampled experiment with the QC pipeline. Considering the TBP
factor ChIP-exo \cite{venters13} and ChIP-nexus \cite{chipnexus}
experiment in human K562 cell lines, we sub-sampled 20M to 50M reads
to cover the sequencing depth's range from all the human experiments
that were considered in this study and evaluated them with QC
pipeline. Figure \ref{fig:eval}C exhibits an increasing trend in the
high-depth adjusted $\mbox{ARC}$ in every panel, where the ChIP-nexus
trend being considerable lower than both ChIP-exo replicates. On the
other hand, in Figure \ref{fig:eval}D the bias remains approximately
constant in ChIP-nexus sub-samples, while it increases for both
ChIP-exo replicates. This suggest that the ChIP-exo replicates have
low library complexity and overall lower quality than the ChIP-nexus
experiment, regardless to the fact that all three experiment are
deeply sequenced with more than 100 million reads each.

\subsection*{High Resolution Binding Site Identification with dPeak and
  ChIP-exo}

\subsubsection*{dPeak outperforms competing methods in discovering closely
  spaced binding events from ChIP-exo and ChIP-Seq data}

After we determined which biases could affect a ChIP-exo experiment's
quality we decided to proceed and examine binding sites in high
resolution on high quality ChIP-exo experiments. We examined first
both aerobic replicates from the \emph{E.Coli} first biosample.

Figures \ref{fig:methods_comp}A and \ref{fig:methods_comp}B compares
binding sites calculated with Peakzilla \cite{peakzilla}, MACE
\cite{mace}, GEM \cite{gem} and dPeak \cite{dpeak} in resolution,
where the resolution is defined as the distance from a RegulonDB
annotation to its closest prediction. For the first replicate (Figure
\ref{fig:methods_comp}A), the resolution calculated with all methods
is comparable, while dPeak outperforms the rest in resolution when
using the second replicate as seen in Figure \ref{fig:methods_comp}B.

Additionally, Figures \ref{fig:methods_comp}C and
\ref{fig:methods_comp}D compares the densities of the parameters
estimated by dPeak for ChIP-exo and SE ChIP-Seq data. As expected, the
$\delta$ parameter which measures the average distance from the
$5\prime$ ends to their respective binding events and the $\sigma$
parameter which is a measure from the reads dispersion around their
respective binding events are lower for ChIP-exo than for SE
ChIP-Seq. Therefore, we can asses that dPeak is accurately
representing that the ChIP-exo fragments are allocated more tightly
around the transcription factor binding sites.

% \color{blue}
% outline

% \begin{enumerate}
% \item \st{Available data, the motivation of this part is two fold:}
%  \begin{itemize}
%  \item \st{we want to list all data sets we obtained}
%  \item \st{we want to emphasize that for the sig70 datasets we have a
%    unique design that allows us to compare chip exo with SE and PE
%    chipseq , the later one for the first time}
%  \end{itemize}
% \item \st{comparison with chip seq} %? with and without dpeak ?}
% \item \st{only in methods. list the align algorithm used}
% \item \st{chip seq qc metrics}
% \item \st{chip exo qc pipeline, description and justification of the steps}
% \begin{itemize}
% \item \st{need to justify the sampling step. for this we may use the depth
%   table and add the factor to normalize one depth to another. explain
%   the confounding of pcr amplification and the reduction of the number
%   of possible mappable positions.}
% \item \st{case study with FoxA1, analysis of enrichment vs library
%   complexity and analysis and the strand imbalance tools. in methods
%   we need to add which tests did we perform to show that the pipeline
%   works. how do we expect from the diagnostics and how to interpret
%   the plots.}
% \item \st{analysis of why the pipeline works. in methods need to specify
%   how and in results need to specify why it works, and what are the
%   conclusions from it}
% \end{itemize}
% \item high resolution analysis with chip exo and comparison with PE
%   chip seq
%   \begin{itemize}
%   \item justify what do we used dpeak (in methods need to list the
%     other methods used). perhaps we can add gof plots in
%     supplement. we also want to compare with the same method,
%     therefore we can use the fact that dpeak have a unified model for
%     PE and SE dpeak
%   \item after comparing with other methods we can compare with PE and
%     SE ChIP-seq, this comparison gives also resolution and sensitivity
%   \item compare samples under same budget of reads
%   \end{itemize} 
% \end{enumerate}

% \color{black}
% Figure \ref{fig:methods_comp} compares the resolution defined as the
% minimum distance between a RegulonDB annotation and a binding site
% predicted by either Peakzilla \cite{peakzilla}, MACE \cite{mace}, GEM
% \cite{gem} or dPeak \cite{dpeak}. In a good dataset such as both of
% the ChIP-exo experiments under aerobic (panel 931 and 933) conditions,
% all the methods are comparable in resolution, and dPeak slightly
% outperforms the rest. On the other hand, for experiments dPeak
% identified binding sites with higher resolution than Mace and
% Peakzilla, but with low resolution than Gem. This may be due that the
% fact that Gem uses sequence information in addition to the aligned
% $5\prime$ end counts that dPeak uses.

\subsubsection*{Systematic comparison of ChIP-Seq vs ChIP-exo under
  varying sequencing depth}

Previously, ChIP-exo and SE ChIP-Seq have been compared only at a
fixed depth level in the literature, while they did not include PE
ChIP-Seq as well either. In order to address this limitation in
previous studies, we sampled a fixed amount of reads for each of the
ChIP-exo, PE ChIP-Seq and SE ChIP-Seq datasets of the $\sig$ samples
($N$ reads for both ChIP-exo experiment and $N/2$ or $N$ pairs for PE
ChIP-Seq to assume more realistic situation of a fixed
cost). Additionally, it is worth noticing that for PE ChIP-Seq we
sampled both ends of the fragment, hence for each sequencing depth we
are sampling the half amount of pairs for PE ChIP-Seq than for
ChIP-exo or SE ChIP-Seq.

% For each sampled dataset we applied our lower-to-higher resolution
% pipeline by calling peaks with MOSAiCS \cite{mosaics} and then
% deconvolving the binding events by using dPeak \cite{dpeak}. For the
% ChIP-exo experiments we called peaks by using the GC-content and
% mappability models with MOSAiCS, since it's background is usually
% composed of scattered reads across the genome; and for the ChIP-Seq
% datasets we used their respective Input samples.

Figure \ref{fig:design} illustrates the behavior of each data type in
$\sig$ experiment under aerobic condition when comparable number of
reads is used for all of ChIP-exo, SE ChIP-Seq and PE ChIP-Seq. In
Figure \ref{fig:design}A we show the number of candidate regions
defined as the number of regions where a binding event was identified
in a collection of high quality peaks; in Figure \ref{fig:design}B we
depict the number of binding events; in Figure \ref{fig:design}C we
show the number of identified targets, where a RegulonDB was
considered identified if a binding event was identified in a 15 bp
vicinity of it; and finally Figure \ref{fig:design}D show the
resolution defined as the distance from a RegulonDB annotation to the
closest dPeak prediction. It is remarkable that even when the number
of candidate peaks or the number of predicted events is lower for
ChIP-exo, it outperforms both PE and SE ChIP-Seq in number of
identified targets and resolution.

% Figure \ref{fig:saturation_rif} shows an analogous analysis but using
% the $\sig$ replicates with and without rif treatment. The left, middle
% and right columns show the fixed depth against the number of predicted
% events, identified targets and resolution being compared at a fixed
% depth level. The behavior of this quantities seems to be opposite to
% the one as in Figure \ref{fig:design}, hence we used the ChIP-exo QC
% pipeline in the fixed depth ChIP-exo experiments. Figure
% \ref{fig:exoQC_sat_aero} shows scatter plots of ARC vs URCR for
% several fixed sample sizes, as the fixed depth increases the two arm
% pattern becomes more distinctive while for lower depth, it seems that
% the majority of the sampled reads were aligned to enriched regions. On
% the other hand in Figures \ref{fig:exoQC_sat1} to
% \ref{fig:exoQC_sat4}, we used the ChIP-exo QC pipeline on the samples
% that are outperformed by PE and SE ChIP-Seq. For a fixed low depth, we
% can see that the majority of the reads are being aligned to
% non-enriched regions since the vertical arm seems to be stronger for
% all 2 conditions and replicates; for higher depth we can see low URCR
% values being predominant, which indicates that the majority of the
% regions being formed by few positions with a higher read
% concentration. In low complexity regions, the reads are being aligned
% to fewer positions but there is no control over the amount of reads
% mapped. Hence, those regions are more likely to being strand-imbalance
% which in turn may bias the binding site estimate and therefore
% decrease the number of identified targets or increase an experiments
% resolution.

\subsubsection*{Comparison with ChIP-Seq data using dPeak}

@@

Figure \ref{fig:reso_all} shows comparisons among ChIP-exo, PE
ChIP-Seq and SE ChIP-Seq. We considered the RegulonDB data as ground
truth, since those are the most recent annotation on \emph{E. Coli}. A
RegulonDB annotation (Salgado et al, 2012\nocite{regulondb}) was
considered to be identified if the distance from the closest dPeak
binding site estimate was less that or equal to $\Sexpr{id_dist}$
bp. That way, the sensitivity is defined as the proportion of
RegulonDB annotations identified in a peak and the resolution is
defined as the minimum distance between a RegulonDB annotation and the
closest dPeak binding site estimate. Figure \ref{fig:reso_all}A shows
that the sensitivity increases as the mean distance between binding
events increases. When the binding events in a peak are closer to each
other, both ChIP-exo and PE ChIP-Seq are comparable, as the distance
increases ChIP-exo identifies a higher proportion of the RegulonDB
annotations; additionally SE ChIP-Seq is significantly less sensitive
than both ChIP-exo and PE ChIP-Seq. Figure \ref{fig:reso_all}B shows
that ChIP-exo and PE ChIP-Seq are comparable in resolution, while both
protocols significantly outperform SE ChIP-Seq.

% When the average distance is greater than 200 bp,
% dPeak identifies more than the 75\% of the binding events in each
% peak, this is intuitive as the mean ChIP-Seq's fragment length is
% shorter than 200 bp, hence the read does not contribute to more than
% one binding event \cite{dpeak}.
% In supplement figures \ref{fig:exoQC_sat1} to \ref{fig:exoQC_sat4}, we
% repeat the same analysis for both replicates under the two rif
% conditions.

@@

  fdr <- 0.01
  topM <- 300
  Ntimes <- prettyNum(1e4,big.mark = ",")
  Msamp <- prettyNum(1e3,big.mark = ",")
  thresh1 <- 10

@@

\textbf{Interpretation of the linear model in the QC pipeline}

The linear model:

\begin{align*}
  \mbox{depth} = \beta_1 \mbox{npos} + \beta_2 \mbox{width}
\end{align*}

is a re-parametrization of the following relationship:

\begin{align}
  \mbox{URCR} = \frac{\kappa}{\mbox{ARC}} + \gamma
\label{mod}
\end{align}

with $\beta_1 = 1 / \gamma$ and $\beta_2 = - \kappa / \gamma$. In this
setting, $\gamma$ can be considered as the large - depth
$\mbox{URCR}$, i.e. the ratio between the number of unique positions
and depth as the depth limits to infinity.

On the other hand, to interpret $\beta_2$, if we assume that the
$\gamma$ value is already given, then by expressing $\kappa$ as a
function of $\mbox{ARC}$ and $\mbox{URCR}$ we can observe the
following identities:

\begin{align*}
  \kappa &= \frac{\mbox{npos}}{\mbox{width}} - \gamma \mbox{ARC} \\
  \frac{\kappa}{\gamma} &= \frac{1}{\gamma} \frac{\mbox{npos}}{\mbox{width}} - \mbox{ARC} 
\end{align*}

where $\mbox{npos}$ and $\mbox{width}$ denote the region's number of
unique position where the reads are aligned to and the region's width
respectively. This is particularly important, $\gamma$ approximates
the true $\mbox{URCR}$ as the sequencing depth increases, which
implies that $- \kappa / \gamma$ can be interpreted as the large
sequencing depth bias of the $\mbox{ARC}$ since as the depth
increases, the first term of $\kappa / \gamma$ is going to approximate
the average read coverage:

\begin{align*}
  \frac{\mbox{npos}}{\mbox{width}} \times \lim_{d \rightarrow
    \infty} \frac{d }{\mbox{npos}} \sim \mbox{ARC}
\end{align*}

Hence, we define the $\mbox{ARC}$ bias as $- \kappa / \gamma$.

@@

\begin{figure}[h!]
  \centering  
  \includegraphics[width = .9\textwidth,page = 1]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \newline
  \includegraphics[width = .9\textwidth,page = 2]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \includegraphics[width = .4\textwidth]{../figs/for_paper/Carroll_FSR_depth_VS_pvalWilcoxon.pdf}
  \caption{Strand imbalance using the mouse FoxA1 experiment from
    \cite{exoillumina}: A) FSR distribution quantiles as the lower
    depth regions are being filtered out, all quantiles approach to
    the median as the lower bound increases. B) Stacked histogram with
    the proportion of regions that are formed by two strands or only
    one, in a good sample the single-stranded regions are going to be
    filtered out quickly as in the middle row. C)
    $-\log_{10}(\text{p.value})$ of testing if the imbalance
    distributions differs when ChIP-exo regions overlap their peaks.}
  \label{fig:strand}
\end{figure}
